Efficient Computation of Extended Surface Sources

### 2024-02-15 repro/papers/GEM22/paper.tex
cleaned up

\title{Efficient Computation of Extended Sources}
\author{William. W. Symes \thanks{The Rice Inversion Project,
Department of Computational and Applied Mathematics, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}

@@

\plot{ddiffprednshsrecpll0}{width=\textwidth}{Plot of difference 
  between data shown in Figures \ref{fig:drecplh0} and 
  \ref{fig:dprednshsrecplh0}, plotted on same color scale as the latter 
  two figures.}
\subsection{Symmetrizing $\Lambda$}

@@

Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
approximate source-to-pressure operator per equation \ref{eqn:winvcomp},
applied once again to the pressure data in Figure
\ref{fig:dsrcphh0}. Note the resemblance to Figures
\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
asymptotic approximations of each other. Figure
\ref{fig:ddiffsymmdxnshsll0} shows the
 the difference between the pressure gather at $z=z_r$ produced from
 the pressure source output by the symmetrized $\Lambda$, and the point source
simulation (Figure \ref{fig:dfwdplh0}), plotted on the same scale as
the latter, in both cases with all propagations in the lens model.

@@

\subsection{Unitary property of modeling operator}
To illustrate this unitary property of $S^+_{z_s,z_r}$, I apply the
operator
\[
  \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
  \Lambda^+_{z_r})
\]
to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecplh0}), in which
$h_s$ is the version (Figure \ref{fig:dinvhslh0}) of the
downgoing source created earlier (Figure \ref{fig:dhslh0}), produced
by inversion of the data in Figure \ref{fig:drecplh0} via the first
equation in display \ref{eqn:approxinv}, propagation in the homogenous
model.  The operator above is computed via the technique explained in
the preceding subsection, below, using
auxiliary receiver arrays 100 m above the data source and receiver arrays.

@@

Figure \ref{fig:compnreslh0} shows the progress of the normal residual
(Euclidean norm of the difference of the two sides of equation \ref{eqn:norm1}),
for Conjugate Gradient and Preconditioned Conjugate Gradient
(Algorithm 1) iterations, applied to solution of the optimization
problem \ref{eqn:einv} with $\alpha=0$. For CG, the norms are both the ordinary
Euclidean norm, $W_m=W_d=I$. For PCG, $W_m$ and $W_d$ are given in
display \ref{eqn:wdef}, with the symmetrized $\Lambda$s computed as
indicated in the preceding subsections. Convergence for the
preconditioned algorithm is roughly 4 times as fast.

\plot{compnres0lh0}{width=\textwidth}{Comparison of normal residual 
  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
  vs. iteration. Data = lens model, point source (Figure 
  \ref{fig:recplh0}), inversion in homogenous model. Penalty weight 
  $\alpha=0$.}

@@

\plot{compnres1lh0}{width=\textwidth}{Comparison of normal residual 
  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
  vs. iteration. Data = lens model, point source (Figure 
  \ref{fig:recplh0}), inversion in homogenous model. Penalty weight 
  $\alpha=10^{-3}$.}

@@

@@

The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
the adjoint state method, in this case a by-product of the
conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awepm}
with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
-
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
\]
\[
+
\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
  p^- + f_r \bf{e_z}) \right)
\]
\[
= 
\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
\]
whence
\begin{equation}
\label{eqn:sadj}
 ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
\end{equation}

The systems \ref{eqn:awepm} for $(p^{\pm},\bv^{\pm})$ with the two
signs of superscript differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
\begin{equation}
  \label{eqn:trdef}
  {\cal R} \left(
    \begin{array}{c}
      p\\
      \bv
    \end{array}
  \right) =
  \left(
    \begin{array}{c}
      Rp\\
      -R\bv
    \end{array}
  \right)
\end{equation}
Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awepm} with
$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
\begin{equation}
  \label{eqn:trsadj}
  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
\end{equation}
Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj}
and \ref{eqn:trsadj} imply equation \ref{eqn:trtr}.

>>>>

\title{Efficient Computation of Extended Surface Sources}
\author{William W. Symes\\
 %\thanks{The Rice Inversion Project,
  Department of Computational and Applied Mathematics,\\
  Rice University,Houston TX 77251-1892 USA,\\
  email {\tt symes@rice.edu},\\
ORCID 0000-0001-6213-4272}

@@

\plot{ddiffprednshsrecpll0}{width=\textwidth}{Plot of difference 
  between data shown in Figures \ref{fig:drecplh0} and 
  \ref{fig:dprednshsrecpll0}, plotted on same color scale as the latter 
  two figures.}
\subsection{Symmetrizing $\Lambda$}

@@

Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
approximate source-to-pressure operator per equation \ref{eqn:winvcomp},
applied once again to the pressure data in Figure
\ref{fig:dsrcphh0}. Note the resemblance to Figures
\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
asymptotic approximations of each other. Figure
\ref{fig:ddiffsymmdnshsll0} shows the
 the difference between the pressure gather at $z=z_r$ produced from
 the pressure source output by the symmetrized $\Lambda$, and the point source
simulation (Figure \ref{fig:dfwdplh0}), plotted on the same scale as
the latter, in both cases with all propagations in the lens model.

@@

\subsection{Unitary property of modeling operator}
To illustrate this unitary property of $S^+_{z_s,z_r}$, I apply the
operator
\[
  \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
  \Lambda^+_{z_r})
\]
to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecplh0}), in which
$h_s$ is the %version (Figure \ref{fig:dinvhslh0}) of the
downgoing source created earlier (Figure \ref{fig:dhshh0})%, produced
%by inversion of the data in Figure \ref{fig:drecplh0} via the first
%equation in display \ref{eqn:approxinv}, propagation in the homogenous
%model.
The operator above is computed via the technique explained in
the preceding subsection, below, using
auxiliary receiver arrays 100 m above the data source and receiver arrays.

@@

Figure \ref{fig:compnres0lh0} shows the progress of the normal residual
(Euclidean norm of the difference of the two sides of equation \ref{eqn:norm1}),
for Conjugate Gradient and Preconditioned Conjugate Gradient
(Algorithm 1) iterations, applied to solution of the optimization
problem \ref{eqn:einv} with $\alpha=0$. For CG, the norms are both the ordinary
Euclidean norm, $W_m=W_d=I$. For PCG, $W_m$ and $W_d$ are given in
display \ref{eqn:wdef}, with the symmetrized $\Lambda$s computed as
indicated in the preceding subsections. Convergence for the
preconditioned algorithm is roughly 4 times as fast.

\plot{compnres0lh0}{width=\textwidth}{Comparison of normal residual 
  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
  vs. iteration. Data = lens model, point source (Figure 
  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
  $\alpha=0$.}

@@

\plot{compnres1lh0}{width=\textwidth}{Comparison of normal residual 
  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
  vs. iteration. Data = lens model, point source (Figure 
  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
  $\alpha=10^{-3}$.}

@@

The linear modeling operator of surface source extended acoustic
waveform inversion is approximately invertible, and this paper has
shown how to approximately invert it. The construction is based on
reverse time propagation of data, as inspired by the literature on
photoacoustic tomography. However, since the input energy comes from a
surface source, rather than a pressure boundary value, the
pressure-to-source operator intervenes. It provides not just an
approximate inverse, but a definition of weighted norms in domain and
range spaces of the modeling operator, in terms of which that operator
is approximately unitary. Accordingly, Krylov space iteration defined
in terms of these weighted norms, or equivalently preconditioned
Conjugate Gradient iteration, gives a rapidly convergent solution
method for the linear subproblem.

The existence of an approximate unitary representation of the modeling
operator is not merely a computational convenience, however. It
reveals fundamental aspects of the operator's structure that enable
an explanation for the mitigation of cycle-skipping, a feature of the
{\em nonlinear} extended inverse problem. This fact echoes earlier observations
concerned a reflected wave inverse problem, involving a modeling
operator with a similar approximate inverse
\cite[]{tenKroode:IPTA14,Symes:IPTA14}. Also, the approximate inverse
leads to a stable computation of the gradient of the nonlinear
objective function \ref{eqn:esi}, resolving a difficulty first noted
also for reflected wave inversion \cite[]{KerSy:94}.

The transmission inverse problem figuring most prominently in
contemporary applied seismology is surely the Full Waveform Inversion
(FWI) of diving wave data. This is essentially the same problem as the
one discussed in this paper, and can be formulated and treated the
same way, at least for acoustic wave physics. All of the topics
treated here are open for elastic wave physics - the analogue of the
pressure-to-source map would is the map from surface velocity field to
corresponding constitutive defect, analogous to the elastic
Dirichlet-to-Neumann map investigated by \cite{Rachele:00}.

The underlying tool in the ideas developed here is geometric optics
(or ray theory), without which the very concept of downgoing waves
would be meaningless. The physics of actual earth materials includes
material heterogeneity on all scales, which appears to leave little
room for the assumption of scale separation underlying geometric
optics. Moreover, earth materials are anelastic, with elastic wave
energy being converted to and from thermal excitation, pore fluid
motion, and so on. A truly satisfactory understanding of inverse wave
problems will eventually need to accommodate heterogenity and
anelasticity beyond the current capabilities of the ray-based theory.

\section{Declarations}
The author received no funding from any source in the performance of
the work reported here, nor does he have any financial or
non-financial interests relevant to the content of this article. The
computations described were developed in the Madagascar reproducible
research framework (www.reproducibility.org). The software source is
available on request from the author.

@@

The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
the adjoint state method, in this case a by-product of the
conservation of energy. This calculation leads to equation
\ref{eqn:sadj1}, from which the other statements about adjoints made in the second
section of the paper follow.

Suppose that $p^-,\bv^-$ solve \ref{eqn:awepm}
with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
-
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
\]
\[
+
\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
  p^- + f_r \bf{e_z}) \right)
\]
\[
= 
\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
\]
whence \ref{eqn:sadj1} follows immediately.
